Tabular Constraint Learning

### 2016-04-10 Paper/ECAI/tabular.tex
filling in the structure and writing the story

\section{Problem Statement}
In the previous section we introduced the problem of tabular constraint learning informally using the example in Figure \ref{fig:main_example}. Here we formalize the statement in terms of Excel constraints and group assignments.

\begin{minipage}[c]{14em}
  \vspace{5pt}
  \begin{tabular}{ll}
    \multicolumn{2}{l}{{\textbf{Tabular Constraint Learning Problem}}}\\
    \vspace{-4pt}
    &\\
    \textbf{Given:}& the set of all groups $\groups$ and of Excel constraints $\constraints$\\
    \textbf{Find:}& for each constraint  $c$ in \constraints all the subgroups of $\groups$\\ 
    & satisfying the \CSignature and \CFunction of $c$ 
  \end{tabular}
\end{minipage}

\section{Approach to Tabular Constraint Learning}
\begin{algorithm}[thb]
  \begin{algorithmic}
    \footnotesize
    \State \textbf{Input:} $\groups$ -- groups, $\constraints$ -- Excel constraints, \dependencies -- dependency graph 
    \State \textbf{Output:} $S$ -- learned constraints with their satisfaction assignments
    \State $S \gets \emptyset$ \Comment{The set of solutions}
  \For{$c \in \constrainttorder(\constraints,\dependencies)$} 
  \State $v_1, ..., v_n =$ variables of $c$ 
    \For{$v_1{:}~G_1, \dots, v_n{:}~G_n \in \generategroups(c, G, S)$}
      \State $S \gets S \cup \findassignment(c, v_1{:}~G_1, \dots, v_n{:}~G_n, S)$
    \EndFor
  \EndFor\\
\Return $S$
\end{algorithmic}
\caption{Tabular constraint learning}
\label{algo:tcl}
\end{algorithm}

Let us describe the key steps and functions in Algorithm \ref{algo:tcl}. Essentially, the algorithm has three steps: constraint ordering, candidate group generation and subgroup satisfaction search, which is in line with the ``generate-and-test'' paradigm that is well-known in AI \cite{whaisasp}. Let us elaborate on each step in detail.

\paragraph{Constraint ordering} $\constrainttorder(\constraints,\dependencies)$ \sergey{explaination that solutions to certain constraints allows us to prune solutions to other constraints -- subsumption, condensed representation}

@@

\subsection{Constraints}
The set of Excel constraints has a partial order in which they should be learned. For certain constraints this order does not matter, such as rank or product. For many others they have to be learned in the order, which is specified as a DAG in Figure \ref{fig:learning_order} (if a constraint is not in the graph, it is independent). This figure specifies a potential parallelization of the computation, since each connected component is independent of the others.

@@

\begin{table*}
  \centering
  \begin{tabular}{lcc}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\
    ALLDIFFERENT(X) & $\textual(G_X) \vee \integer(G_X) $ & $X \in G_X{:}~\forall i,j \in \nat{:}~i \neq j \rightarrow X[i] \neq X[j]$ \\
    X = Y & & \\
    FK -> PK & & \\
    R = PRODUCT(FV, FK=PK | PV) & & \\
    FV = FUZZY-LOOKUP(FK, PK, PV) & & \\
    FV = LOOKUP(FK, PK, PV) & & \\
    PERMUTATION(X) & & \\
    R = O1 * O2 & & \\
    R = PROJECT(P) & & \\
     Y = RANK(X)     & $\numeric(G_X) \wedge \integer(G_Y) \wedge \length(G_X) = \length(G_Y)$ & $\begin{array}{r} X \in G_X, Y \in G_Y{:}~\forall i,j \in \nat{:}~(Y_i \leq Y_j \rightarrow X_{Y_i} \leq X_{Y_j}) \\ \wedge~ 1 \leq Y_i \leq \length(G_Y) \end{array}$ \\
    A = PREV(A) + P - N & & \\
    SERIES(X) & & \\
    Y = SUM(X, col) & & \\
    Y = SUM(X, row) & & \\
    R = SUMIF(FK=PK, V) & & \\
    R = SUMPRODUCT(O1, O2) & & \\

@@

@@

@@

We conclude that for this example our primary goal of finding all constraints is achieved.
A number of redundant constraints ($33.33\%$ of the total) are also found, which, as we will show, is rather high compared to other spreadsheets.
However, this example also contains many short vectors which increases the chance for constraints to be true by accident.

>>>>

\section{Problem Statement}
In the previous section we introduced the problem of tabular constraint learning informally using the example in Figure \ref{fig:main_example}. Here we formalize the statement in terms of Excel constraints and group assignmentsd as follows:

\begin{minipage}[c]{14em}
  \vspace{5pt}
  \begin{tabular}{ll}
    \multicolumn{2}{l}{{\textbf{Tabular Constraint Learning Problem}}}\\
    \vspace{-4pt}
    &\\
    \textbf{Given:}& the set of all groups $\groups$ and of Excel constraints $\constraints$\\
    \textbf{Find:}& for each constraint  $c$ in \constraints all the subgroups of $\groups$\\ 
    & satisfying the \CSignature and \CFunction of $c$ \\
  \end{tabular}
  \vspace{6pt}
\end{minipage}

  The key observation here is that essentially the problem is a constraint enumeration problem, where each constraint is independent. This property comes from fact that in Excel each formula is applied based on the existing data in the cells. This allows learning constraints independently of each other by examining constraint satisfaction on the groups.

\section{Approach to Tabular Constraint Learning}
\begin{algorithm}[thb]
  \begin{algorithmic}
    \footnotesize
    \State \textbf{Input:} $\groups$ -- groups, $\constraints$ -- Excel constraints, \dependencies -- dependency graph 
    \State \textbf{Output:} $S$ -- learned constraints with their satisfaction assignments
    \State $S \gets \emptyset$ \Comment{The set of solutions}
  \For{$c \in \constrainttorder(\constraints,\dependencies)$} 
  \State $v_1, ..., v_n =$ variables of $c$ 
    \For{$v_1{:}~G_1, \dots, v_n{:}~G_n \in \generategroups(c, G, S, \dependencies)$}
      \State $S \gets S \cup \findassignment(c, v_1{:}~G_1, \dots, v_n{:}~G_n, S)$
    \EndFor
  \EndFor\\
\Return $S$
\end{algorithmic}
\caption{Tabular constraint learning \sergey{Samuel, I think \generategroups takes \dependencies as an argument}}
\label{algo:tcl}
\end{algorithm}
In theory each constraint is independent and equally useful. In practice, however, things are different. Assume that for some $X$ and $Y$ it holds that \textit{SUM(X)=Y}, and also we discovered that the vector $Y$ is equal to a vector $Z$. Then, of course, \textit{SUM(X)=Z} holds and it is a valid solution to the sum constraint problem. From a user's perspective the last solution is useless, since he or she already knows that $Y=Z$ and the latter constraint satisfaction follows. To capture this condensed representation \cite{condensed} of solutions, we have introduced a subsumption (generality) order between constraints, which is depicted using \textcolor{red}{dashed} lines in Figure \ref{fig:learning_order}. Besides generality, we know that certain constraints have as a subproblem other constraints, e.g. if $X$ is a \textit{series}, then it is also a \textit{permutation}. If discovered that $X$ is a not permutation, then $X$ should not be tested for the series constraint at all. To capture this solving strategy we introduced a constraint construction order depicted using \textcolor{green}{solid} lines in Figure \ref{fig:learning_order}.

Essentially, our Algorithm \ref{algo:tcl} has three steps: constraint ordering, candidate group generation and subgroup satisfaction search, which is in line with the ``generate-and-test'' paradigm that is well-known in AI \cite{whaisasp}. Let us elaborate on each step in detail.

\paragraph{Constraint ordering} $\constrainttorder(\constraints,\dependencies)$ uses the DAG \dependencies in Figure \ref{fig:learning_order} in the following way: all constraints are generated in the topological order of \dependencies. Also, if there is a construction edge from $C_1$ to $C_2$, and $X$ is not a solution to $C_1$, then $X$ is not generated as a candidate group for $C_2$. If there is a sumsumption edge from $C_1$ to $C_2$ and $X$ is a solution to $C_1$, then $X$ is not generated as a candidate group for $C_2$.

@@

\subsection{Constraints}
As emphasized before, the set of Excel constraints has a partial order in which they should be learned. For certain constraints this order does not matter, such as \textit{rank} or \textit{product}. For many others they have to be learned in the order, which is specified as a DAG in Figure \ref{fig:learning_order} (if a constraint is not in the graph, it is independent). This figure also specifies a potential parallelization of the computation, since each connected component is independent of the others.

\sergey{need to elaborate here especially on the structure and how the constraints we picked}

@@

\begin{table*}
  \centering
  \begin{tabular}{lcc}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\
    ALLDIFFERENT(X) & & \\
    X = Y & & \\
    FK $\rightarrow$ PK & & \\
    R = PRODUCT(FV, FK=PK | PV) & & \\
    FV = FUZZY-LOOKUP(FK, PK, PV) & & \\
    FV = LOOKUP(FK, PK, PV) & & \\
    PERMUTATION(X) & & \\
    R = O1 * O2 & & \\
    R = PROJECT(P) & & \\
     Y = RANK(X)   & & \\
    A = PREV(A) + P - N & & \\
    SERIES(X) & & \\
    Y = SUM(X, col) & & \\
    Y = SUM(X, row) & & \\
    R = SUMIF(FK=PK, V) & & \\
    R = SUMPRODUCT(O1, O2) & & \\

@@

\section{Declarative modeling}
\sergey{We need to fit ASP, Minizinc and all that here, I mean it is supposed to be an important point after all}

@@

\sergey{I think there should be three things: key example evaluated, benchmark of 30 spreadsheets and maybe the FBI example, how we can detect interesting dependecies and correct potential mistakes}

@@

We conclude that for this example our primary goal of finding all constraints is achieved.
A number of redundant constraints ($\frac{1}{3}$ of the total) are also found, which, as we will show, is rather high compared to other spreadsheets.
However, this example also contains many short vectors which increases the chance for constraints to be true by accident.
